---
title: "BayesCNA minimal example"
output:
  html_document:
    df_print: paged
    highlight: pygments
    theme: readable
---

# BayesCNA

In this script, we give an example output of BayesCNA so that the user can validate that the program is running as it should. For more details of the method, we refer the reader to our preprint: <https://www.biorxiv.org/content/10.1101/2025.09.22.677749v1>

## Installation of packages and settings

Below, we provide installation instructions for the BCP package. If BCP is already installed on your machine, this step can be skipped.

```{r}
library(devtools) # version 2.4.5
devtools::install_github("swang87/bcp")
```

The packages below are required to run BayesCNA and import its functions. Version numbers used are provided as comments.

```{r}
# load required packages
library(ggplot2)    # version 3.5.1
library(magrittr)   # version 2.0.3
library(bcp)        # version 4.0.3
library(pracma)     # version 2.4.4

# load BayesCNA functions
source("../segmentation.R")
```

## Loading the data

For illustration purposes, we have simulated an artificial CNA profile with added Gaussian noise, with purity$p = 0.15$ and $n = 1000$ genomic bins. This provides us with a ground truth for evaluation. The simulated data contains the true copy number profile (scaled by the purity) and the observed (noisy) copy number profiles:

```{r}
example.data <- readRDS("example_data.RData")
true.CNAs <- as.numeric(example.data$segment) # true copy number profile
observed.CNAs <- as.numeric(example.data$CN) # observed (noisy) copy number profile
```

Below we plot the data. The red signal indicates the true copy number profile.

```{r}
plot.simulated.data <- \(dataset) {
  N.features <- length(dataset$CN)
  data.frame(
    signal = c(dataset$CN),
    index = rep(1:N.features)
  ) %>% ggplot(aes(x = index, y = signal)) + geom_line(color = "lightgray") +
    geom_line(data = data.frame(index = 1:N.features, signal = dataset$segment %>% as.numeric), color = "firebrick") +
    labs(x = "Genomic bin", y = "Observed CNAs")
}

plot.simulated.data(example.data)
```

BayesCNA works with *any user provided binned genomic data*. However, a key assumption of the BCP model is constant Gaussian noise so it is recommended to process the data (e.g correct for GC and mappability bias) to not violate this assumption. To run BayesCNA on user provided data from this script, exchange the variable observed.CNAs to contain your data.

## Running BayesCNA

The key parameter of the Bayesian changepoint detection (BCP) model is the parameter $p_0$, that is the prior on changepoint probabilities. A small $p_0$ is suggested if we know a priori that that the sample is of high coverage and/or purity, as this generally increases the precision, and a larger $p_0$ is suggested if the sample is of low coverage and/or purity, i.e. is noisy. We choose 500 burnin iterations and 2000 Markov Chain Monte Carlo (MCMC) samples. The burnin samples are not used for estimating the posterior changepoint probabilities.

```{r}
# BCP specific parameters
p <- 0.01
n.burnin <- 500
n.mcmc <- 2000 
# Run BCP
denoised <- bcp(observed.CNAs, p0 = p, burnin = n.burnin, mcmc = n.mcmc, return.mcmc = T)
```

Below is the posterior means and changepoints probabilities, provided by the BCP-package. High posterior changepoint probabilities indicate likely positions of CNAs in the genome. High probabilities are assigned to prominent CNAs.

```{r}
plot(denoised)
```

The next step is to identify a likely partition of the genome based on the posterior changepoint probabilities. Important parameters for the post-processing are $\varepsilon$ (percentage of the highest peak), that governs how much noise will be removed from posterior changepoint probabilities and $\eta$, that governs how much merging of segments will be performed.

```{r}
eps <- 0.1; eta <- 0.5
denoised.postprocessed <- bcp.pp.sd(denoised$data[, 2], denoised, eps = eps, n.sd = eta) 
cat("Identified breakpoints: ", denoised.postprocessed)
```

```{r}
plot.simulated.data(example.data) + geom_vline(xintercept = denoised.postprocessed, color = "black", linetype = "dashed")
```

From the identified breakpoints, we construct the copy number profile by computing the median of the data, conditioned on the obtained partition:

```{r}
cna.profile <- construct.profile(observed.CNAs, denoised.postprocessed, genome.length = 1000)
```

We can also run the entire BayesCNA pipeline directly in the following way

```{r}
results <- run.BayesCNA(observed.CNAs, p = p, eps = eps, eta = eta, n.mcmc = n.mcmc, n.burnin = n.burnin)
```

The output of the run.BayesCNA function is the identified changepoints and the reconstructed signal.
